### 2017-05-18 01-intro.tex
Gone over figures etc. Still need to redo related work.
Polish a bit on GPU plots

The problem of detecting overlapping communities is modeled by the mixed membership stochastic blockmodels (MMSB)~\cite{airoldi2009mixed} and in this paper we are particularly interested in a variant of MMSB, called assortative MMSB (a-MMSB\footnote{Although we work on a-MMSB for simplicity, it is also straightforward to apply the proposed method to the general MMSB model.}) \cite{gopalan2012scalable}.
The MMSB model is a probabilistic graphical model~\cite{koller2009probabilistic} that represents a convenient paradigm for modeling complex relationships between a potentially large number of random variables. Bayesian graphical models, where we define priors and infer posteriors over parameters, also allow us to quantify model uncertainty and facilitate model selection and averaging. However, an increasingly urgent question is whether these models and their inference procedures will be up to the challenge of handling very large graphs.

@@

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms, like deep learning, in several
ways. First, it is highly data-intensive which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
communities.

>>>>

The problem of detecting overlapping communities is modeled by the mixed membership stochastic block model (MMSB)~\cite{airoldi2009mixed} and in this paper we are particularly interested in a variant of MMSB, called assortative MMSB (a-MMSB\footnote{Although we work on a-MMSB for simplicity, it is also straightforward to apply the proposed method to the general MMSB model.}) \cite{gopalan2012scalable}.
The MMSB model is a probabilistic graphical model~\cite{koller2009probabilistic} that represents a convenient paradigm for modeling complex relationships between a potentially large number of random variables. Bayesian graphical models, where we define priors and infer posteriors over parameters, also allow us to quantify model uncertainty and facilitate model selection and averaging. However, an increasingly urgent question is whether these models and their inference procedures will be up to the challenge of handling very large graphs.

@@

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms, like deep learning, in several
ways. First, it is highly data-intensive which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
community graphs.

### 2017-05-18 03-related.tex
Polish a bit on GPU plots

\textbf{\textit{\large Need to rewrite to better fit the Background Section}}

Immediately when GPUs evolved into generally programmable accelerators, machine
learners recognized the potential of GPUs for their science.
Much research
has been devoted to deep-learning implementations on GPUs. However, Mixed
Membership deduction (MMSB) is usually considered untractable for deep-learning
algorithms. In contrast, our algorithm belongs to the family of Bayesian
(approximative, stochastic) methods, where efficient algorithms for large-scale
data have recently been developed. A broad survey of Bayesian methods is given
in~\cite{DBLP:journals/corr/Zhu0H14}. Broadly speaking, Bayesian approximative
algorithms are subdivided into variational and Monte Carlo methods; Markov
Chain Monte Carlo (MCMC) is a subgroup of the latter. GPUs have been put to
use in MCMC algorithms, e.g.\ \cite{journals/bioinformatics/MedlarGSBK13}
uses GPUs to analyse parental linkage patterns in a biology context and
\cite{DBLP:journals/csda/WhiteP14} does the same to model terrorist activity.
Variational inference on MMSB is described
in~\cite{DBLP:conf/nips/GopalanMGFB12}, but they don't use GPUs. Latent
Dirichlet Allocation, another variety of Bayesian Approximation, is used on
GPUs by~\cite{DBLP:conf/nips/YanXQ09}.

Still, for high-dimension problems, the standard MCMC algorithms fall short in
performance. Recently, varieties of MCMC algorithms have been developed where
gradient information is used to speed up convergence. Langevin and Hamiltonian
dynamics are representatives of these
varieties~\cite{Girolami_riemannmanifold}.  Our algorithm uses Riemann Manifold
Langevin dynamics. In \cite{beam2014fast}, GPUs are used to perform Hamiltonian
descent, using Python interfaces to access the standard cuBLAS
library~\cite{cuBLAS}. They limit their optimizations for the GPU to
saving on data transfers between host and device memory. They do not apply
their framework to MMSB.

A different MMSB algorithm with stochastic gradient descent on the GPU is
the Online Tensor approach~\cite{DBLP:journals/corr/HuangNHVA13}. They deduce
overlapping communities for many of the same datasets as this paper, and they
report fast convergence. Their implementation uses the cuBLAS
library, and, unlike our work, there is no attempt to hand-optimize the GPU
kernels. Additionally, the algorithmic approach to solve the problem is
fundamentally different.
Since they target GPUs only, the datasets they
can handle are limited by the device memory of the GPU. Our implementation converges
quickly on the GPU but can also function with reduced speedup on a multicore
CPU. In the latter case, the dataset
size is only limited by the memory of the host machine.

@@

There are a few projects that did a distributed implementation
for community detection on very large network graphs. In contrast to our work,
they all detect non-overlapping communities: \cite{Bu2013246} investigate a
large number of networks; \cite{2015arXiv150302115L} investigate hierarchical
stochastic blockmodels; \cite{Prat-Perez:2014:HQS:2566486.2568010} use
multi-core machines.

\begin{comment}
\subsection{Related work on RDMA Key-Value stores}
In this paper, we describe our custom RDMA D-KV (Distributed Key-Value)
store. Current RMDA D-KV store implementations are RamCloud~\cite{RamCloud},
Pilaf~\cite{Pilaf}, Herd~\cite{Herd} and FaRM~\cite{FaRM}. All these systems
use RDMA to implement a D-KV store. However, all of them are far more powerful
than our custom implementation -- and this power comes at a cost that we
can avoid. They implement a generic D-KV store that controls concurrency,
supports dynamic inserts and deletes, supports variable-sized values
(whose size may change at an update), and keys of arbitrary type. Because
of the nature of our distributed algorithm, we have to deal with none of
these issues. For us, values are fixed-size, allocated only at the initial
population, and remain alive forever. We have no concurrency between writes
and reads or other writes. Our keys are a contiguous range of integers. All
these properties together allow an extremely low-overhead implementation
that does not involve the remote host in any transaction.
\end{comment}

>>>>

We consider related work for both parts of this paper: acceleration and scaling
of our MCMC-MMSB algorithm. Acceleration of deep learning algorithms on GPUs
is an ongoing success story. Although approximative Baysian algorithms are
not such a natural fit for GPUs, a number of projects explores this terrain.
E.g.~\cite{journals/bioinformatics/MedlarGSBK13} uses GPUs with their
MCMC approach to analyse parental linkage patterns in a biology context and
\cite{DBLP:journals/csda/WhiteP14} does the same to model terrorist activity.
Latent Dirichlet Allocation, another variety of Bayesian Approximation,
is used on GPUs by~\cite{DBLP:conf/nips/YanXQ09}. There is also related
work on the variants of MCMC algorithms that use the gradient to speed
up convergence. Langevin and Hamiltonian dynamics are representatives of
these varieties~\cite{Girolami_riemannmanifold}. Our algorithm uses Riemann
Manifold Langevin dynamics. In \cite{beam2014fast}, GPUs are used to perform
Hamiltonian descent, using Python interfaces to access the standard cuBLAS
library~\cite{cuBLAS}. They limit their optimizations for the GPU to saving
on data transfers between host and device memory. They do not apply their
framework to MMSB.

A different MMSB algorithm with stochastic gradient descent on the GPU is
the Online Tensor approach~\cite{DBLP:journals/corr/HuangNHVA13}. They deduce
overlapping communities for many of the same datasets as this paper, and they
report fast convergence. Their implementation uses the cuBLAS
library, and, unlike our work, there is no attempt to hand-optimize the GPU
kernels. Additionally, the algorithmic approach to solve the problem is
fundamentally different.
Since they target GPUs only, the datasets they
can handle are limited by the device memory of the GPU. Our implementation converges
quickly on the GPU but can also use multicore CPUs. In the latter case, speedup is
limited but the larger memory allows tackling larger datasets.

\cite{liakos-bigdata16} also addresses overlapping community detection, but their
algorithm uses a local dispersion-aware approach. They find communities for 5000
nodes in each of the large SNAP datasets, see Table~\ref{table-snap}.

@@

There are a few projects that implemented a distributed algorithm
for community detection on very large network graphs. In contrast to our work,
they all detect non-overlapping communities: \cite{Bu2013246} investigate a
large number of networks; \cite{2015arXiv150302115L} investigate hierarchical
stochastic blockmodels; \cite{Prat-Perez:2014:HQS:2566486.2568010} use
multi-core machines. The results of our distributed MCMC-aMMSB implementation
have been previously published at the 2016 ParLearning
workshop~\cite{DBLP:conf/ipps/El-HelwHLAWB16}, where it earned a best-paper
award.

\textbf{\emph{QUESTION: Re-insert related work on RDMA KV stores?}}
\begin{comment}
\subsection{Related work on RDMA Key-Value stores}
In this paper, we describe our custom RDMA D-KV (Distributed Key-Value)
store. Current RMDA D-KV store implementations are RamCloud~\cite{RamCloud},
Pilaf~\cite{Pilaf}, Herd~\cite{Herd} and FaRM~\cite{FaRM}. All these systems
use RDMA to implement a D-KV store. However, all of them are far more powerful
than our custom implementation -- and this power comes at a cost that we
can avoid. They implement a generic D-KV store that controls concurrency,
supports dynamic inserts and deletes, supports variable-sized values
(whose size may change at an update), and keys of arbitrary type. Because
of the nature of our distributed algorithm, we have to deal with none of
these issues. For us, values are fixed-size, allocated only at the initial
population, and remain alive forever. We have no concurrency between writes
and reads or other writes. Our keys are a contiguous range of integers. All
these properties together allow an extremely low-overhead implementation
that does not involve the remote host in any transaction.
\end{comment}

### 2017-05-18 distr-03-design.tex
Polish a bit on GPU plots

%\looseness=-1
$\pi$~and $\phi$ are 32-bit float arrays of size $K \times N$. For our
largest distributed experiment, com-Friendster with $N$=64M and $K$=12K, each
requires~3TB. We decided to store only $\pi$ and $\sum\phi$, and recalculate
$\phi$ from these whenever appropriate. Here, we trade a substantial gain
in memory usage against some computation. $\pi$ is partitioned across the workers,
and $\pi$ values are accessed via a DKV (distributed key-value) store.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 7 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi+\sum\phi$
are written to the DKV store.

\textit{update\_beta} (lines 10 and 11 in Algorithm~\ref{algorithm}) requires $|\Minibatch| \times K$ operations. It is preceded by an MPI
barrier to ensure that up-to-date $\pi$ values are read. \textit{update\_beta}
is split into four steps. The first stage partitions the mini-batch across
machines to calculate contributions to $g_{ab}(\theta)$. The values of $\pi$ for
the local mini-batch vertices are loaded from the DKV store. These calculations
are done with thread parallelism over the mini-batch vertices. In the second step,
a multi-threaded summation is performed to calculate the contributions per
machine, followed by a distributed summation using an MPI reduce operation. The
third step calculates $\beta$ from $\theta$ sequentially at the master. This
operation takes only $K$ steps. The resulting $\beta$ is broadcast with MPI
to the workers.

>>>>

%\looseness=-1
$\pi$~and $\phi$ are 32-bit float arrays of size $K \times N$. For our
largest distributed experiment, com-Friendster with $N$=64M and $K$=12K, each
requires~3TB. Analogously to the single-machine optimization described in
Section~\ref{sec-gpu}, we store only $\pi$ and
$\sum\phi$, and recalculate
$\phi$ from these whenever appropriate. Here, we trade a substantial gain
in memory usage against some computation. $\pi$ is partitioned across the workers,
and $\pi$ values are accessed via a DKV (distributed key-value) store.

@@

%\looseness=-1
These updates are used by \textit{update\_pi} (line 9 in
Algorithm~\ref{algorithm}) locally, besides the value of
$\pi/\phi$ for the mini-batch vertices. The update to $\pi$ requires
$M \times K$ operations, and it is done in parallel over mini-batch vertices. Because of
memory consistency, this stage awaits completion of \textit{update\_phi} with
an MPI barrier. After the calculation, the updated values of $\pi,~\sum\phi$
are written to the DKV store.

\textit{update\_beta} (lines 12 and 15 in Algorithm~\ref{algorithm}) requires $|\Minibatch| \times K$ operations. It is preceded by an MPI
barrier to ensure that up-to-date $\pi$ values are read. \textit{update\_beta}
is split into four steps. The first stage partitions the mini-batch across
machines to calculate contributions to $g_{ab}(\theta)$. The values of $\pi$ for
the local mini-batch vertices are loaded from the DKV store. These calculations
are done with thread parallelism over the mini-batch vertices. In the second step,
a multi-threaded summation is performed to calculate the contributions per
machine, followed by a distributed summation using an MPI reduce operation. The
third step calculates $\beta$ from $\theta$ sequentially at the master. This
operation takes only $K$ steps. The resulting $\beta$ is broadcast with MPI
to the workers.

### 2017-05-18 distr-04-evaluation.tex
Gone over figures etc. Still need to redo related work.
Polish a bit on GPU plots

This section presents an in-depth empirical analysis of the design discussed in
Section~\ref{design-section}. To this end, we assess the scalability and
efficiency of the solution using multiple criteria.
%
First, the \textit{weak scaling} of the solution is assessed with respect to increasing
numbers of targeted latent communities.
%
Second, an evaluation of the \textit{strong scaling} behavior of the distributed
algorithm is presented.
%
Third, the effects of \textit{pipelining the computation} and varying the number of latent communities on the system's
performance are analyzed.
%
Next, we \textit{contrast} the effectiveness of scaling the computation horizontally
and vertically.
%
Further, we evaluate the efficiency and overhead associated with \textit{network
communication} between cluster nodes.
%
Finally, we apply the algorithm to large data sets and comment on its execution time.

Empirical results for the distributed implementation were obtained from experiments on the VU and
Leiden University
DAS5 clusters which consist of 68 and 24 compute nodes respectively. Each
compute node is equipped with a dual 8-core Intel Xeon E5-2630v3 CPU clocked
at 2.40GHz, 64GB of memory and 8TB of storage. Moreover, the compute nodes of
each site are interconnected by
FDR~InfiniBand. The MPI implementation used is OpenMPI, which supports
user-space Infiniband~\cite{gabriel04:_open_mpi}.

We compared the distributed performance against a shared-memory virtual machine
at SURFsara's HPC Cloud system with 40~cores and 1TB of memory; its underlying
physical system contains 40 Intel Xeon E7-4850 cores clocked at 2.00GHz, and
it does not oversubscribe resources.

@@

\begin{figure}[t] % [htbp]
  \centering
  \epsfig{file=plots/sweep-over-K-proportional-np.eps, width=\columnwidth}
  \caption{Average execution time per algorithm iteration when varying
  the number of latent communities proportionally to the number of compute
  nodes.}
  \label{fig-weak-scaling}
\end{figure}

@@

\begin{figure}[t] % [htb]
  \centering
  \epsfig{file=plots/sweep-over-np-fixed-K.eps, width=\columnwidth}
  \caption{(a)~Execution time of 2048 algorithm iterations for the
  same problem size (com-Friendster, K=1024, M=16384 n=32) across different
  cluster sizes. (b)~Speedup achieved for same experiments in (a) with respect to
  8 nodes}
  \label{fig-strong-scaling}
\end{figure}

@@

\begin{table}[t] % [htbp]
  \centering
  \begin{tabular}{l d{5.1} d{5.1}}
    \textbf{Iteration (sub)stages}  & \multicolumn{1}{c}{non-pipelined}
                                          & \multicolumn{1}{c}{pipelined} \\
    \hline
    draw/deploy mini-batch     &  45.6  &  26.2 \\
    load $\pi$                 & 205    & 209 \\
    update $\phi$              &  74    &  74 \\
    \hline
    total                      & 331    & 241 \\
    \hline
    \\[0.2ex]
  \end{tabular}
  \caption{Stages that profit from the two pipelining optimizations.
  com-Friendster on 65
  compute nodes, with 12K communities. Times in ms per iteration.}
  \label{table-pipeline}
\end{table}

@@

\begin{figure}[t] % [htb]
  \centering
  \epsfig{file=plots/sweep-over-K-fixed-np.eps, width=\columnwidth}
  \caption{Performance effect of varying the number of communities on the
    algorithm's execution time on 64 nodes when disabling or enabling both
    pipelining optimizations.}
  \label{fig-pipeline}
\end{figure}

@@

%################################################
\subsubsection{Horizontal vs. Vertical Scalability}
\begin{figure}[t] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\columnwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on one machine with 40~cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
considerably simpler approach is developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we used the virtual
machine with 40~cores and 1TB of
memory at SURFSara's HPC Cloud system. By provisioning all 40~cores
we ensured that there is no
resource contention from other users. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16~cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up}-b tests the performance of the HPC Cloud
system compared to 64 nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

\begin{figure}[t] % [htbp]
  \centering
  \epsfig{file=plots/qperf.eps, width=\columnwidth}
  \caption{Performance of DKV store vs.\ \textit{qperf} and MPI roundtrip.}
  \label{fig-qperf}
\end{figure}

>>>>

This section presents an in-depth empirical analysis of our distributed design.
To this end, we assess the scalability and
efficiency of the solution using multiple criteria.
%
First, the \textit{weak scaling} of the solution is assessed with respect to increasing
numbers of targeted latent communities.
%
Second, an evaluation of the \textit{strong scaling} behavior of the distributed
algorithm is presented.
%
Third, the effects of \textit{pipelining the computation} and varying the number of latent communities on the system's
performance are analyzed.
%
Next, we \textit{contrast} the effectiveness of scaling the computation horizontally
and vertically.
%
Further, we evaluate the efficiency and overhead associated with \textit{network
communication} between cluster nodes.
%
Finally, we apply the algorithm to large data sets and comment on its execution time.

As noted in Section~\ref{sec-experimental}, experimental results
were obtained using the DAS5 clusters.
The MPI implementation used is OpenMPI, which supports
user-space Infiniband~\cite{gabriel04:_open_mpi}.

Moreover, we compared the distributed performance against a shared-memory virtual machine
at SURFsara's HPC Cloud system with 40~cores and 1TB of memory; its underlying
physical system contains 40 Intel Xeon E7-4850 cores clocked at 2.00GHz, and
it does not oversubscribe resources.

@@

\begin{figure}[tb] % [htbp]
  \centering
  \epsfig{file=plots/sweep-over-K-proportional-np.eps, width=\columnwidth}
  \caption{Average execution time per algorithm iteration when varying
  the number of latent communities proportionally to the number of compute
  nodes.}
  \label{fig-weak-scaling}
\end{figure}

@@

\begin{figure}[tb] % [htb]
  \centering
  \epsfig{file=plots/sweep-over-np-fixed-K.eps, width=\columnwidth}
  \caption{(a)~Execution time of 2048 algorithm iterations for the
  same problem size (com-Friendster, K=1024, M=16384 n=32) across different
  cluster sizes. (b)~Speedup achieved for same experiments in (a) with respect to
  8 nodes}
  \label{fig-strong-scaling}
\end{figure}

@@

\begin{table}[tb] % [htbp]
  \centering
  \begin{tabular}{l d{5.1} d{5.1}}
    \textbf{Iteration (sub)stages}  & \multicolumn{1}{c}{non-pipelined}
                                          & \multicolumn{1}{c}{pipelined} \\
    \hline
    draw/deploy mini-batch     &  45.6  &  26.2 \\
    load $\pi$                 & 205    & 209 \\
    update $\phi$              &  74    &  74 \\
    \hline
    total                      & 331    & 241 \\
    \hline
    \\[0.2ex]
  \end{tabular}
  \caption{Stages that profit from the two pipelining optimizations.
  com-Friendster on 65
  compute nodes, with 12K communities. Times in ms per iteration.}
  \label{table-pipeline}
\end{table}

@@

\begin{figure}[tb] % [htb]
  \centering
  \epsfig{file=plots/sweep-over-K-fixed-np.eps, width=\columnwidth}
  \caption{Performance effect of varying the number of communities on the
    algorithm's execution time on 64 nodes when disabling or enabling both
    pipelining optimizations.}
  \label{fig-pipeline}
\end{figure}

@@

%################################################
\subsubsection{Horizontal vs. Vertical Scalability}
\begin{figure}[tb] % [htb]
  \centering
  \epsfig{file=plots/hpc-cloud.eps, width=\columnwidth}
  \caption{Performance comparison between the distributed implementation
  running on DAS5 and the multi-threaded solution on one machine with 40~cores and
  1TB of RAM.}
  \label{fig-scale-up}
\end{figure}
One of the main drawbacks of designing a distributed solution for a given
algorithm is the inherent complexity of communication and synchronization. A
considerably simpler approach is developing a multi-threaded version and
running it on a machine with abundant memory and CPU cores. In such a context,
access to all of the algorithm's state would be an order of magnitude faster
than RDMA. Additionally, the overhead associated with synchronizing threads is
negligible compared to using MPI primitives. To evaluate the efficacy of both
approaches we used the virtual
machine with 40~cores and 1TB of
memory at SURFSara's HPC Cloud system. By provisioning all 40~cores
we ensured that there is no
resource contention from other users. Figure~\ref{fig-scale-up}
reports the execution time per algorithm iteration for two experimental setups.
%
First, Figure~\ref{fig-scale-up}-a shows the performance of executing the
algorithm on the HPC Cloud
system with 40 and 16 cores compared to a single DAS5 node with 16~cores. This
test uses the com-DBLP data set to enable the use of a large number of
communities without running out of memory. It is clear from the results that
the performance can benefit from the additional cores provided by HPC Cloud system.
%
Further, Figure~\ref{fig-scale-up}-b tests the performance of the HPC Cloud
system compared to 64 nodes of the DAS5
cluster using the com-Friendster data set. Clearly, the parallel and
distributed implementation vastly
outperforms the single-node multi-threaded solution. Moreover, the trajectory
of both curves shows a widening gap between them suggesting that the relative
performance difference will increase for larger $K$.
%
In conclusion, the overhead of network communication in the distributed version
is more than compensated by the increasing compute power compared to
a single-node implementation.

@@

\begin{figure}[tb] % [htbp]
  \centering
  \epsfig{file=plots/qperf.eps, width=\columnwidth}
  \caption{Performance of DKV store vs.\ \textit{qperf} and MPI roundtrip.}
  \label{fig-qperf}
\end{figure}

### 2017-05-18 gpu-04-02-sequential.tex
Polish a bit on GPU plots

\subsection{An Efficient Sequential Baseline Version}
\label{sec-design-sequential}

Initially, the algorithm was implemented in Python~\cite{LiAW15}, as is common
with machine learning researchers and practitioners. It
relied on numpy~\cite{numpy} to perform the numerical computations without the
overhead associated with Python's interpreter. However, the algorithm also
depended on native Python data structures such as sets, dictionaries and
lists, which are far less efficient.
This section presents how the Python implementation was
converted into an efficient sequential C++ application which serves as a
baseline for the parallel implementations.

Initially, the Python code base was ported into equivalent C++ code.
The program structure remained unchanged and the Python data structures were
simply substituted with their corresponding alternatives in C++.

>>>>

The MCMC-aMMSB algorithm that is the subject of this paper was initially
implemented in Python~\cite{LiAW15}, as is common
with machine learning researchers and practitioners. It
relied on numpy~\cite{numpy} to perform the numerical computations without the
overhead associated with Python's interpreter. However, the algorithm also
depended on Python data structures such as sets, dictionaries and
lists, which are far less efficient.
This section describes how the Python implementation was
converted into an efficient sequential C++ application which serves as a
baseline for the parallel and distributed implementations.

First, the Python code base was ported into equivalent C++ code.
The program structure remained unchanged and the Python data structures were
simply substituted with their corresponding alternatives in C++.

### 2017-05-18 gpu-04-03-accelerator.tex
Polish a bit on GPU plots

\subsection{Restructuring for Parallelism}

@@

% Naive strategy
\paragraph*{\textbf{The naive strategy}} simply allocate temporary vectors in thread
local memory to store $Probs$ and $Grads$ which physically resides in
device memory. All memory accesses are coalesced to achieve the highest
possible bandwidth to and from device memory.

% Shared memory caching strategy
\paragraph*{\textbf{The shared memory strategy}} allocate the temporary vectors
$Probs_i$ and $Grads_i$
in shared memory. Furthermore, this strategy copies the $\pi_i$ of the selected
mini-batch vertex to shared memory in order to avoid repeated reads of the same
memory locations in device memory

% Register and shared memory using code generation and loop unrolling 
\paragraph*{\textbf{The code generation strategy}} dynamically generate the code of the kernel to
custom tailor its properties. For example, this strategy can control whether a
vector should be placed or cached in shared memory. Additionally, it can
control which vectors explicitly reside in registers by allocating space on the
stack frame, unrolling all inner loops of the kernel and substituting all
vector addressing with static values. The code generation strategy allows this
flexibility for the three vectors of concern, namely, $Probs_i$, $Grads_i$
and~$\pi_i$.
Hence, this strategy allows for 8 possible configurations denoted by
three letters each of which is a choice between $R$ or $S$ to represent
$Register$ or $Shared$ respectively. For example, $SSR$ denotes that $Probs_i$,
$Grads_i$ and $\pi_i$ were placed in $Shared$, $Shared$ and $Register$
respectively.

>>>>

% \section{Accelerating on one Many-core Machine}

@@

% Naive strategy
\paragraph*{\textbf{Naive strategy}} simply allocate temporary vectors in thread
local memory to store $Probs$ and $Grads$ which physically reside in
device memory. All memory accesses are coalesced to achieve the highest
possible bandwidth to and from device memory.

% Shared memory caching strategy
\paragraph*{\textbf{Shared memory strategy}} allocate the temporary vectors
$Probs_i$ and $Grads_i$
in shared memory. Furthermore, this strategy copies the $\pi_i$ of the selected
mini-batch vertex to shared memory in order to avoid repeated reads of the same
memory locations in device memory

% Register and shared memory using code generation and loop unrolling 
\paragraph*{\textbf{Code generation strategy}} dynamically generate the code of the kernel to
custom tailor its properties. For example, this strategy can control whether a
vector should be placed or cached in shared memory. Additionally, it can
control which vectors explicitly reside in registers by allocating space on the
stack frame, unrolling all inner loops of the kernel and substituting all
vector addressing with static values. The code generation strategy allows this
flexibility for the three vectors of concern, namely, $Probs_i$, $Grads_i$
and~$\pi_i$.
Hence, this strategy allows for 8 possible configurations denoted by
three letters each of which is a choice between $R$ or $S$ to represent
$Register$ or $Shared$ respectively. For example, $SSR$ denotes that $Probs_i$,
$Grads_i$ and $\pi_i$ were placed in $Shared$, $Shared$ and $Register$
respectively.

### 2017-05-18 gpu-04-design.tex
Place the chunks of text unmodified into one document
Start working on integration: block out double algorithm, unify symbol usage
Some nitpicks
Gone over figures etc. Still need to redo related work.
Polish a bit on GPU plots

\section{System Design and Implementation}

The process of designing an accelerated version of the SG-MCMC MMSB algorithm
involved multiple transformations. First, an efficient C++ baseline version
was created. Further, changes to the algorithm and the data structures
were carried out, for both efficient resource utilization and parallelization.
This section provides an overview of the system's evolution in incremental
phases, identifying the key contributions and differences between consecutive
states.

\input{04-02-sequential}

\input{04-03-accelerator}

>>>>

% \subsection{System Design and Implementation}

Based on the efficient C++ baseline version,
changes to the algorithm and the data structures
were carried out, for both efficient resource utilization and parallelization.
This section provides an overview of the system's evolution in incremental
phases, identifying the key contributions and differences between consecutive
states. This section concludes with an in-depth evaluation of the performance
of our various optimizations, on multi-core CPUs as well as a number of GPU
accelerators.

% \input{gpu-04-02-sequential}

\input{gpu-04-03-accelerator}

### 2017-05-18 gpu-05-02-sequential.tex
Gone over figures etc. Still need to redo related work.
Polish a bit on GPU plots

\subsubsection{Achieving an Efficient Baseline Sequential Version}

@@

Due to the performance limitations of the Python implementation, the
experimental configuration of this section is minimal.
%
Namely, we use the CA-HepPh dataset; the number of
iterations is 1000.
%Note that these trial runs stop very far before convergence; using a small
%number of
%iterations is valid because the time spent in an iteration does not vary
%during the lifetime of the algorithm.
The selected number of communities is~1024, the mini-batch size~32, the
neighbor sample size~32.

@@

\begin{table}[b]
\center\begin{tabular}{l d{5.1} d{2.1} d{2.1} d{2.1}}
	&	& \multicolumn{3}{c}{C++} \\
\cline{3-5}
	& & 
            \multicolumn{1}{c}{Python} &
                \multicolumn{1}{c}{float64} &
		    \multicolumn{1}{c}{float32} \\
\multicolumn{1}{l}{Algorithm stage} &\multicolumn{1}{c}{Python} &
            \multicolumn{1}{c}{idiom} &
                \multicolumn{1}{c}{(baseline)} &
		    \multicolumn{1}{c}{baseline} \\
\hline
sample mini-batch       &      0.03 &  0.03 &  0.03 & 0.02 \\
sample neighbor sets    &      8.5  &  0.5  &  0.4  & 0.35 \\
update\_phi             &  9,014    & 52.7  &  8.9  & 4.9  \\
update\_pi              &     93.1  & 40.3  &  0.11 & 0.05 \\
update\_theta\_beta     &    339    &  2.1  &  0.76 & 0.55 \\
perplexity*             &    178    &  0.18 &  0.18 & 0.26 \\
\hline
\\[-1ex]
\end{tabular}
\caption{Performance comparison between Python, C++ that follows the Python
idioms, and our baseline C++ implementations. 1000 iterations, times in
\textrm{ms} per iteration; *perplexity time divided by 100 iterations.}
\label{table-python-comparison}
\end{table}

@@

>>>>

\subsection{Performance of Sequential Versions}

@@

Due to the performance limitations of the Python implementation, the
experimental configuration of this section is minimal.
%
Namely, we use the small CA-HepPh dataset from the SNAP graph collection, see
Table~\ref{table-snap}; the number of
iterations is 1000.
%Note that these trial runs stop very far before convergence; using a small
%number of
%iterations is valid because the time spent in an iteration does not vary
%during the lifetime of the algorithm.
The selected number of communities is~1024, the mini-batch size~32, the
neighbor sample size~32.

@@

\begin{table}[tb]
\center\begin{tabular}{l d{5.1} d{2.1} d{2.1} d{2.1}}
	&	& \multicolumn{3}{c}{C++} \\
\cline{3-5}
	& & 
            \multicolumn{1}{c}{Python} &
                \multicolumn{1}{c}{float64} &
		    \multicolumn{1}{c}{float32} \\
\multicolumn{1}{l}{Algorithm stage} &\multicolumn{1}{c}{Python} &
            \multicolumn{1}{c}{idiom} &
                \multicolumn{1}{c}{(baseline)} &
		    \multicolumn{1}{c}{baseline} \\
\hline
sample mini-batch       &      0.03 &  0.03 &  0.03 & 0.02 \\
sample neighbor sets    &      8.5  &  0.5  &  0.4  & 0.35 \\
update\_phi             &  9,014    & 52.7  &  8.9  & 4.9  \\
update\_pi              &     93.1  & 40.3  &  0.11 & 0.05 \\
update\_theta\_beta     &    339    &  2.1  &  0.76 & 0.55 \\
perplexity*             &    178    &  0.18 &  0.18 & 0.26 \\
\hline
\\[-1ex]
\end{tabular}
\caption{Performance comparison between Python; C++ that follows the Python
idioms; and our baseline C++ implementations. 1000 iterations, times in
\textrm{ms} per iteration; *perplexity time divided by 100 iterations.}
\label{table-python-comparison}
\end{table}

@@

A first multi-threaded version was created by annotating the
baseline implementation with OpenMP primitives in a straightforward
fashion. This accelerated the stages \textit{update\_phi} and
\textit{update\_pi} with a factor~4 for the very small problem instance
in Table~\ref{table-python-comparison}. For larger problem instances, the
speedup for these dominant phases grew to a factor~10. The vectorizing OpenCL
backend described in the next Section again significantly improves the
performance, but at the cost of substantial coding effort.
\textbf{\emph{Question: speak about the other stages? they are pipelined.}}
